Adopting the Euro: a Synthetic Control Approach \\ Online Appendix\\ \vspace{2em}

### 2024-01-28 Text/Manuscript.tex
Updated text

\title{Adopting the Euro: a Synthetic Control Approach\footnote{We thank Benjamin Born, David Hope, Evi Pappa, Donghai Zhang, Laura Puzello, Martin Kornejew, Moritz Schularick, Pedro Gomis-Porqueras, Vitor Possebom as well as Paola Boel and the remaining research division members at the Sveriges Riksbank for their feedback and discussions. We likewise thank seminar participants at the University of Bonn and participants of the $13^{th}$ Annual Meeting of the Portuguese Economic Journal. We would like to thank the Sveriges Riksbank for hosting us while conducting part of the research that led to this paper. We acknowledge financial support from the Fundação para a Ciência e Tecnologia [projects references SFRH/BD/144581/2019 and SFRH/BD/144820/2019] the Deutsche Forschungsgemeinschaft (German Research Foundation) under Germany's Excellence Strategy [EXC 2126/1 - 390838866] and under the RTG 2281 - The Macroeconomics of Inequality.}}

\author{\hspace{-2em} Ricardo Duque Gabriel\footnote{Corresponding author: University of Bonn and NBER. E-mail address: ricardofilipeduquegabriel@gmail.com} \hspace{5em} Ana Sofia Pessoa\footnote{University of Bonn. E-mail address: sofiamendespessoa@gmail.com}\\ \hspace{-3em} University of Bonn and NBER \hspace{3em} University of Bonn}
%\date{December, 2020}

\begin{document}
\begin{titlepage}
\clearpage\thispagestyle{empty}
\maketitle
\thispagestyle{empty}
\vspace{-3em}
\begin{abstract}
We investigate whether joining the European Monetary Union and losing the ability to set monetary policy affected the economic growth of Eurozone countries. We use the synthetic control approach to create a counterfactual scenario for how each Eurozone country would have evolved without adopting the euro. We let this matching algorithm determine which combination of other developed economies best resembles the pre-euro path of twelve Eurozone economies. Our estimates suggest that there were some mild losers (France, Germany, Italy, and Portugal) and a clear winner (Ireland). Nevertheless, a GDP decomposition suggests that the drivers of the economic gains and losses are heterogeneous. In particular, our results show that adopting the euro spurred government consumption and deterred private consumption. The common currency also stimulated trade in most cases but only Germany, Luxembourg, and Ireland bear positive net trade benefits.
\end{abstract}

@@

In January 1999, the exchange rates between member countries’ national currencies and the euro were fixed irrevocably and the European Central Bank (ECB) officially took over the responsibility of conducting the unified monetary policy. Twenty years have passed since the euro was launched and the member states gave up the ability to set their own monetary policy. In this work, we evaluate whether joining the euro had any macroeconomic effect for twelve of the Eurozone countries and investigate which were the channels driving this effect.

@@

Our contribution is two-folded. First, we evaluate the macroeconomic impact of adopting the euro measured by the real gross domestic product (GDP). Theoretical predictions about this effect are ambiguous and depend on whether the costs outweigh the benefits of joining the Eurozone.\footnote{We refer the reader to \cite{Lane2006} and \cite{Beetsma2010} who provide a more recent account of the real effects of the European Monetary Union by surveying the literature on its macroeconomic costs and benefits.} Indeed, we find that there are some mild losers (France, Germany, Italy, and Portugal) and a clear winner (Ireland). Such heterogeneous findings are in line with  \cite{DeGrauwe2020} and \cite{Puzzello2018}.

Most notably, we perform a GDP decomposition exercise to investigate which were the channels driving the output gains and losses and whether they differed from country to country. For Ireland, the private consumption and investment notably explain almost 80\% of the total output gain from joining the euro. While for France and Portugal, the private consumption and the net exports accounted for a large share of the economic loss, in the case of Germany and Italy, the private consumption and the investment explain the negative impact of the Euro. For most countries, the trade volume was significantly higher than if they had not joined the Eurozone. Nonetheless, the common currency had a positive impact on the trade balance solely for Germany and Ireland.

%\textbf{Literature Review:} 
This paper is related to several strands of the literature. The first one is directly related to the methodology used to construct the counterfactuals. To employ the SCM, we follow the original work by \cite{Abadie2003} and  \cite{Abadie2010, Abadie2015} who developed the methodology.\footnote{A good overview of the literature using this methodology can be found in \cite{Abadie2019}.} Furthermore, we follow more recent work by \cite{Campos2018} who evaluate the impact of the European Union accession, \cite{Born2019} who assess the macroeconomic impact of the election of Donald Trump as the President of the USA, and finally \cite{Breinlich2020} and \cite{Born2018} who study the costs of economic nationalism by looking at the Brexit vote impact on the transactions and GDP, respectively.

The second strand relates to the political economy of monetary unions. \cite{Feldstein1997}

@@

More recent work from \cite{Alesina2002} explains that forgoing monetary policy, on the one hand, implies losing a stabilization device to deal with domestic shocks but, on the other hand, can boost credibility and price stability. \cite{Alesina2002} show that if there is a reduction in trading costs, the adoption of a common currency has a direct positive effect on trade, output, and consumption.\footnote{\cite{Alesina2002B} and \cite{Barro2007} empirically investigate and present evidence in accordance with these theoretical predictions.}

There is also a broad literature that tests empirically these theoretical links. At about the time the euro was launched, \cite{Rose2000} famously estimated that a currency union could boost up to three times bilateral trade. The relevance of these results to the euro case was immediately doubted since the sample used for the analysis was based on unions of small, poor, and remote countries. \cite{Micco2003} developed the first comprehensive study for the impact of the European Monetary Union (EMU) on trade, concluding that the euro had a positive impact not only on trade between member states but also with third parties. Some papers have also highlighted the impact of currency unions on investment. Among others, \cite{Barr2003} suggest that inward investment in the countries outside the union would have been greater if they had joined the EMU. Furthermore, \cite{DeSousa2011} estimate that, in the Eurozone countries, investment increased with the single currency adoption.

@@

The weights are chosen to minimize the difference between each treated unit and its doppelganger's pre-intervention outcome variable and predictors. We follow Born et al. (\citeyear{Born2018}, \citeyear{Born2019}) and define the outcome variable as real GDP which is normalized to unity in 1970 in each country.\footnote{The date of normalization corresponds to the beginning of the sample as explained in Section \ref{SS_Data}. Moreover, focusing on the normalized per capita real GDP instead does not qualitatively change the results.}

@@

Secondly, since the counterfactual weights are constructed according to the pre-intervention characteristics, we have to assure that there are no (external) differentiated shocks during the study period in the donor pool countries \citep{Abadie2019}. To account for this issue, contrarily to \cite{Fernandez2015} and \cite{Verstegen2017}, we only consider observations until 2007. From 2008 onward, the Great Recession affected countries in very different ways and arguably provoked structural changes in the affected economies.

It is also important to exclude any country that was treated from the donor pool. In this context, this is addressed by using only donor countries that never adopted the Euro. One important consideration on countries belonging to the European Union (EU) must be made: even though the euro adoption and the EU accession are intertwined, the cumulative gains of entering the EU for Denmark and the UK stopped before 1999 and are insignificant for Sweden --- according to Figure 1 in \cite{Campos2018}. Hence, we add EU countries that did not adopt the euro to the donor pool.

Furthermore, policy interventions frequently have spillover effects to non-treated units. When employing the SCM, it is important to ensure that the counterfactuals are not affected by the treatment. In our analysis, this is equivalent to ruling out the possibility that the euro adoption by an individual country affected the outcome variable of the donor countries. This assumption is tested by performing in-space placebo testes in section \ref{SS_spaceplacebo}.

Fourth, the intervention has no effect on the outcome before the implementation period. In section \ref{SS_timeplacebo}, potential anticipation effects are tested by changing the treatment date used in the analysis.

@@

It is expected that the SCM yields an imperfect pre-treatment match for some countries given that our procedure determines 14 parameters (country weights) to match 37 observations. Notwithstanding, this methodology can provide substantial improvement relative to alternative methods as differences-in-differences \citep{Ferman2019} and thus, we are confident that this data-driven approach is the best to study the problem at hands.

Table \ref{TA_weights} displays the donor country weights that constitute each doppelganger. For instance, the synthetic Spain is composed by all countries in the donor pool yet being significantly constructed by using data from the United States (46\%), Mexico (19\%), Switzerland (18\%), and Australia (16\%). We are overall confident on the plausibility and credibility of the methodology weighting scheme.\footnote{Potential concerns regarding the use of countries which belong to the Exchange Rate Mechanism (Denmark, Sweden, and the United Kingdom) are addressed in Section \ref{SS_donor}.}

@@

\begin{figure}[h!]
    \centering
    \caption{The Impact of the Eurozone Accession}
    \includegraphics[scale=1.15]{SCM_gdp_Annual.eps}
    \annote{In each graph, the dashed line represents the normalized real GDP for the synthetic country and the continuous line represents the series for the actual country. The shaded area corresponds to two standard deviations of the difference between the treated country and the doppelganger prior to the euro accession. The vertical line represents the treatment period - 1999 for all countries except for Greece which is 2001. For each country, the analysis starts in 1970 and ends in 2007. %The baseline specification has as predictors labor productivity growth, employment share, and GDP shares from PENN Tables 9.1 data. OECD restricted donor pool.
    }
    \label{F_1}
\end{figure}

@@

\begin{table}[h!]
\scriptsize
\caption{\label{TA_Gap} Doppelganger Output Gap}\centering
\begin{tabular}{lcccccccccccc} 
\toprule
 & \textbf{AUT}  & \textbf{BEL}  & \textbf{FIN}  & \textbf{FRA}  & \textbf{DEU}  & \textbf{GRC}  & \textbf{IRL}  & \textbf{ITA}  & \textbf{LUX}  & \textbf{NLD}  & \textbf{PRT}  & \textbf{ESP} \\
 \midrule  
\textbf{Gap} & \multirow{2}{*}{-1,712} & \multirow{2}{*}{-1,668} & \multirow{2}{*}{777} & \multirow{2}{*}{-2,632} & \multirow{2}{*}{-5,788} & \multirow{2}{*}{-1,098} & \multirow{2}{*}{10,781} & \multirow{2}{*}{-6,089} & \multirow{2}{*}{4,697} & \multirow{2}{*}{168} & \multirow{2}{*}{-2,558} & \multirow{2}{*}{900} \\
(\texteuro \ per capita) &  &  &  &  &  & &  & &  & &  &  \\
\bottomrule
\end{tabular}
\annote{This table presents the doppelganger output gap per capita in 2007. This measure is obtained by adjusting the real GDP gap for the population size and converting 2011 US dollars into 2011 euro. We use the conversion rate available from the PWT 9.1 for this year ($\approx 0.73$).}
\end{table}

@@

A key assumption to study the impact of a policy intervention is that there is no reverse causality. In our context, this means that countries must not have adopted the euro due to economic considerations. This assumption is plausible because the Eurozone accession was driven mainly by political rather than economic factors \citep{Eichengreen1993, Feldstein1997}. In fact, by not satisfying the requirements of an Optimum Currency Area, many economists believed that countries adopting the euro would face economic losses \citep{Jonung2009}. This argument holds even for the Greek case which had decided to join the euro before the single currency was a reality.\footnote{According to the 1998 convergence report from the European Commission, Greece did not join the single currency in 1999 because it had not fulfilled any of the four convergence criteria. Notwithstanding, the decision of joining was already made.}

Sections \ref{SS_Implementation} and \ref{SS_Data} discussed the conditions under which the SCM provides suitable estimates of causal effects and this section addresses some of these requirements. To further back the notion that the doppelganger gap is indeed caused by the euro adoption, Sections \ref{SS_timeplacebo} and \ref{SS_spaceplacebo} provides two key placebo experiments. We can be confident that the synthetic control estimator captures the causal effect of an intervention as long as similar magnitudes are not estimated in cases where the intervention did not take place \cite{Born2018}. Finally, Section discusses the statistical significance of the results \ref{SS_stat}.

@@

\begin{figure}[h!]
    \centering
    \caption{In-time Placebo Tests}
    \includegraphics[scale=1.2]{Maastricht.eps}
    \annote{In each graph, the blue dashed line represents the normalized real GDP for the synthetic country and the black full line represents the series for the actual country. The shaded area corresponds to two standard deviations of the difference between the treated country and the doppelganger prior to the euro accession. The vertical line depicts the placebo treatment period - 1992 for all countries. For all countries, the analysis starts in 1970 and ends in 2007. %The baseline specification has as predictors labor productivity growth, employment share, and GDP shares from PENN Tables 9.1 data. OECD restricted donor pool.
    }
    \label{F_Maastricht}
\end{figure}

@@

The absence of anticipation effects for the remaining countries may be due to two things. First, the key event representing a change for most European citizens was the irrevocably the exchange rate fix on 31st December 1998 and the euro launch on the 1st of January 1999. Second, most of these countries had already experienced trade and economic gains from joining the European Union \citep{Campos2018}. Therefore, such effects lie in our pre-treatment sample and thus, are already being considered.
%Still, these figures should be analysed carefully. For example, the plot for Greece shows some effect that starts even before the placebo treatment of 1992. This anticipated effect seems to be driven by the Greek European Union (EU) accession in 1981 and not from the Euro. 
%Furthermore, estimates for Austria and Finland must be interpreted carefully as these countries entered the EU exactly in 1995 and the placebo treatment effect may be biased.

@@

Second, there might have been spillover effects of the treatment onto the donor countries. We address this issue by iteratively re-estimating our baseline estimate for each Eurozone country excluding in each iteration one of the countries with positive weight \citep{Abadie2015}.

We display this robustness check for countries that weighted, at least, 10\% in the construction of, at least, 2 countries' counterfactuals in Appendix \ref{SS_donor_appendix}. This exercise shows that no particular donor country is driving the main conclusions. So, it is unlikely that neither spillover effects nor one specific country in the donor pool is driving the results.
%\footnote{For the sake of brevity, the figures for all the remaining countries are not reported, but they can be provided by the authors upon request. The results remain qualitatively unchanged.} So, it is unlikely that spillover effects nor one specific country in the donor pool are driving the results.

@@

For countries which joined the EU at least 10 years before the Eurozone creation, we re-do the estimates using their accession data as the start of our sample, that is, Ireland (1973), Greece (1981), Portugal and Spain (1986). From Figure \ref{F_Start}, it is possible to conclude that the results from the baseline analysis are robust.

@@

\section{What Drives the Doppelganger Gap? \label{S_Components}}

In this section, we take one step further and investigate what drives the results presented in Figure \ref{F_1} by decomposing the euro accession response of GDP into the response of its components. First, we compute the series for each GDP component for both countries and corresponding counterfactuals. Then, we try to understand what explains the output gains and losses from the accession by accounting the role of each component for the doppelganger GDP gap.

@@

@@

For most countries, government consumption is one of the components with a positive contribution. This is a particularly puzzling finding given the Eurozone countries were operating under the fiscal rules of the Stability and Growth Pact during the post-treatment period. Nevertheless, according to \cite{Bofinger2003}, the correlation between the average real growth rate of government consumption and fiscal deficits between 1999 and 2002 was actually negative meaning that the low deficit rule does not necessarily implies less government spending. Our results, obtained using a more careful econometric application during the sample period of 1999-2007, provide evidence in favor of the later hypothesis.

@@

Table \ref{T_Decomposition_NX} indeed reveals that, with the exception of France and Italy, all countries had a higher trade volume than if they had not adopted the common currency. This result is in accordance with \cite{Baldwin2008} and \cite{Schmitz2011} who argue that the euro has significantly promoted trade in the Eurozone countries. Yet, net exports changed differently across countries. Only Germany and Ireland experienced significant net trade benefits from the euro accession. We corroborate the negative impact on net exports in most cases as documented in \cite{Hope2016}.

@@

\section{Conclusion \label{S_Conclusion}}
In this paper we study the impact of the EMU accession on the macroeconomic performance of the first twelve member states. We use the synthetic control method to construct a counterfactual of these countries' GDP. This method allows building a doppelganger which should represent the economic activity of these countries in the absence of the euro adoption.

Our findings suggest that there are mild losers (France, Germany, Italy, and Portugal) and a clear winner (Ireland). Notwithstanding, the drivers of such estimates are heterogeneous as our GDP decomposition analysis indicates. Our results show that for the majority of Eurozone countries, euro spurred government consumption and deterred investment and private consumption. The common currency also stimulated trade for most cases but only Germany and Ireland bear positive net trade benefits.

>>>>

\title{Adopting the Euro: a Synthetic Control Approach\footnote{We thank Benjamin Born, David Hope, Evi Pappa, Donghai Zhang, Laura Puzello, Martin Kornejew, Moritz Schularick, Pedro Gomis-Porqueras, Vitor Possebom as well as Paola Boel and the remaining research division members at the Sveriges Riksbank for their feedback and discussions. We likewise thank seminar participants at the University of Bonn and participants of the $13^{th}$ Annual Meeting of the Portuguese Economic Journal. We would like to thank the Sveriges Riksbank for hosting us while conducting part of the research that led to this paper. We acknowledge financial support from the Fundação para a Ciência e Tecnologia [projects references SFRH/BD/144581/2019 and SFRH/BD/144820/2019] the Deutsche Forschungsgemeinschaft (German Research Foundation) under Germany's Excellence Strategy [EXC 2126/1 - 390838866] and under the RTG 2281 - The Macroeconomics of Inequality. The views expressed herein are those of the author and should not be attributed to the Federal Reserve Board, or those affiliated with the Federal Reserve System.}}

\author{\hspace{-2em} Ricardo Duque Gabriel\footnote{Corresponding author: Federal Reserve Board. E-mail address: ricardofilipeduquegabriel@gmail.com} \hspace{5em} Ana Sofia Pessoa\footnote{University of Bonn. E-mail address: sofiamendespessoa@gmail.com}\\ \hspace{-1em} Federal Reserve Board \hspace{5em} University of Bonn}
%\date{December, 2020}

\begin{document}
\begin{titlepage}
\clearpage\thispagestyle{empty}
\maketitle
\thispagestyle{empty}
\vspace{-3em}
\begin{abstract}
We investigate whether joining the European Monetary Union and losing the ability to set monetary policy affected the economic growth of Eurozone countries. We use the synthetic control approach to create a counterfactual scenario for how each Eurozone country would have evolved without adopting the euro. We let this matching algorithm determine which combination of other developed economies best resembles the pre-euro path of twelve Eurozone economies. Our estimates suggest that most countries' economic growth was not significantly affected. There were some mild losers (France, Germany, Italy, and Portugal) and a clear winner (Ireland). Nevertheless, the drivers of the economic gains and losses are heterogeneous. First, we find Ireland's economic gains are more modest when excluding profits and income earned by foreigners. Second, our results show that adopting the euro spurred government consumption and deterred private consumption. The common currency also stimulated trade in most cases but only Luxembourg bears positive net trade benefits.
\end{abstract}

@@

In January 1999, the exchange rates between member countries’ national currencies and the euro were fixed irrevocably and the European Central Bank (ECB) officially took over the responsibility of conducting the unified monetary policy. Twenty-five years have passed since the euro was launched and the member states gave up their monetary policy independence. A single currency could potentially boost economic growth by eliminating exchange rate risks, facilitating flows of goods, people, and capital, and promoting competition and economic integration of the member states. In this work, we evaluate whether joining the euro had any macroeconomic effect for twelve of the Eurozone countries and investigate which were the channels driving this effect.

@@

Our contribution is two-fold. First, we evaluate the macroeconomic impact of adopting the euro measured by the real gross domestic product (GDP). Theoretical predictions about this effect are ambiguous and depend on whether the costs outweigh the benefits of joining the Eurozone.\footnote{We refer the reader to \cite{Lane2006} and \cite{Beetsma2010} who provide a more recent account of the real effects of the European Monetary Union by surveying the literature on its macroeconomic costs and benefits.} Indeed, we find that there are some mild losers (France, Germany, Italy, and Portugal) and a clear winner (Ireland). Such heterogeneous findings are in line with  \cite{DeGrauwe2020} and \cite{Puzzello2018}.

Most notably, we perform a GDP decomposition exercise to investigate which channels drove the output gains and losses and whether they differed from country to country. For Ireland, private consumption and investment notably explain almost 80\% of the total output gain from joining the euro. While for France and Portugal, private consumption and net exports accounted for a large share of the economic loss, in the case of Germany and Italy, private consumption and investment explain the negative impact of the Euro. For most countries, the trade volume was significantly higher than if they had not joined the Eurozone. Nonetheless, the common currency had a positive impact on the trade balance solely for Germany and Ireland.

%\textbf{Literature Review:} 
\textbf{Literature Review.} This paper is related to several strands of the literature. The first one is directly related to the methodology used to construct the counterfactuals. To employ the SCM, we follow the original work by \cite{Abadie2003} and  \cite{Abadie2010, Abadie2015} who developed the methodology.\footnote{A good overview of the literature using this methodology can be found in \cite{Abadie2019}.} Furthermore, we follow more recent work by \cite{Campos2018} who evaluate the impact of the European Union accession, \cite{Born2019} who assess the macroeconomic impact of the election of Donald Trump as the President of the USA, and finally \cite{Breinlich2020} and \cite{Born2018} who study the costs of economic nationalism by looking at the Brexit vote impact on the transactions and GDP, respectively.

The second strand of the literature relates to the political economy of monetary unions. We build on work by \cite{Eichengreen1993,  Feldstein1997, Dyson1999, Willett2000, James2012}; and  \cite{Spolaore2013} who find that countries adopted the Euro because of political rather than economic motivations, to rule out reverse causality and study the macroeconomic impact of joining the Euro.

@@

More recent work from \cite{Alesina2002} explains that forgoing monetary policy, on the one hand, implies losing a stabilization device to deal with domestic shocks but, on the other hand, can boost credibility and price stability. \cite{Alesina2002} show that if there is a reduction in trading costs, the adoption of a common currency has a direct positive effect on trade, output, and consumption.\footnote{\cite{Alesina2002B} and \cite{Barro2007} empirically investigate and present evidence of these theoretical predictions.}

There is also a broad literature that empirically tests these theoretical links. At about the time the euro was launched, \cite{Rose2000} famously estimated that a currency union could boost up to three times bilateral trade. The relevance of these results to the euro case was immediately doubted since the sample used for the analysis was based on unions of small, poor, and remote countries. \cite{Micco2003} developed the first comprehensive study for the impact of the European Monetary Union (EMU) on trade, concluding that the euro had a positive impact not only on trade between member states but also with third parties. Some papers have also highlighted the impact of currency unions on investment. Among others, \cite{Barr2003} suggest that inward investment in the countries outside the union would have been greater if they had joined the EMU. Furthermore, \cite{DeSousa2011} estimate that, in the Eurozone countries, investment increased with the single currency adoption.

@@

The weights are chosen to minimize the difference between each treated unit and its doppelganger's pre-intervention outcome variable and predictors.

Aiming to study the macroeconomic impact of the Euro accession, we use real GDP as the main outcome variable of this analysis. First, real GDP is one of the most important metrics to assess countries' degree of real convergence and economic growth. Second, it allows us to unpack the baseline result by studying how each GDP component changed after the Euro accession. \footnote{Using GDP per capita as the outcome variable yields similar results. See Figure \ref{F_normgdppc} in the appendix.} Following Born et al. (\citeyear{Born2018}, \citeyear{Born2019}), we normalize real GDP to unity in the first sample year, 1970. \footnote{Focusing on the normalized per capita real GDP instead does not qualitatively change the results.}

@@

Secondly, since the counterfactual weights are constructed according to the pre-intervention characteristics, we have to ensure that there are no (external) differentiated shocks during the study period in the donor pool countries \citep{Abadie2019}. To account for this, we only consider observations until 2007. From 2008 onward, the great financial recession and the European sovereign debt crisis affected countries in very different ways and might have caused structural changes in the affected economies. While before the great recession, there were no major differences in average economic growth, after 2008 treated units grew on average much slower than the donor pool countries.

It is also important to exclude any country that was treated from the donor pool. In this context, this is addressed by using only donor countries that never adopted the Euro. Yet, Denmark, Sweden, and the UK, which belonged to the European Union (EU) but did not adopt the Euro, are included in the donor sample. One important consideration is that, even though the euro adoption and the EU accession are intertwined, the cumulative gains of entering the EU stopped before 1999 for Denmark and the UK and are insignificant for Sweden according to \cite{Campos2018}.

Furthermore, policy interventions frequently have spillover effects on non-treated units. When employing the SCM, it is important to ensure that the counterfactuals are not affected by the treatment. In our analysis, this is equivalent to ruling out the possibility that the euro adoption by an individual country affected the outcome variable of the donor countries. This assumption is tested by performing in-space placebo tests in section \ref{SS_spaceplacebo}.

Fourth, the intervention does not affect the outcome before the implementation period. In section \ref{SS_timeplacebo}, potential anticipation effects are tested by changing the treatment date used in the analysis.

@@

It is expected that the SCM yields an imperfect pre-treatment match for some countries given that our procedure determines 14 parameters (country weights) to match 37 observations. Notwithstanding, this methodology can provide substantial improvement relative to alternative methods as differences-in-differences \citep{Ferman2019}, and thus, we are confident that this data-driven approach is the best to study the problem at hand.

Table \ref{TA_weights} displays the donor country weights that constitute each doppelganger. For instance, the synthetic Spain is composed by all countries in the donor pool, and yet it is significantly constructed by using data from the United States (46\%), Mexico (19\%), Switzerland (18\%), and Australia (16\%). We are overall confident on the plausibility and credibility of the methodology weighting scheme.\footnote{Potential concerns regarding the use of countries that belong to the Exchange Rate Mechanism (Denmark, Sweden, and the United Kingdom) are addressed in Section \ref{SS_donor}.}

@@

\begin{figure}[h!]
    \centering
    \caption{The Impact of the Eurozone Accession}
    \includegraphics[scale=1.15]{SCM_gdp_Annual.pdf}
    \annote{In each graph, the dashed line represents the normalized real GDP for the synthetic country and the continuous line represents the series for the actual country. The shaded area corresponds to two standard deviations of the difference between the treated country and the doppelganger prior to the euro accession. The vertical line represents the treatment period - 1999 for all countries except for Greece which is 2001. For each country, the analysis starts in 1970 and ends in 2007. %The baseline specification has as predictors labor productivity growth, employment share, and GDP shares from PENN Tables 9.1 data. OECD restricted donor pool.
    }
    \label{F_1}
\end{figure}

@@

\begin{table}[h!]
\scriptsize
\caption{\label{TA_Gap} Doppelganger Output Gap}\centering
\begin{tabular}{lcccccccccccc} 
\toprule
 & \textbf{AUT}  & \textbf{BEL}  & \textbf{FIN}  & \textbf{FRA}  & \textbf{DEU}  & \textbf{GRC}  & \textbf{IRL}  & \textbf{ITA}  & \textbf{LUX}  & \textbf{NLD}  & \textbf{PRT}  & \textbf{ESP} \\
 \midrule  
\textbf{Gap} & \multirow{2}{*}{-682} & \multirow{2}{*}{-678} & \multirow{2}{*}{1,269} & \multirow{2}{*}{-2,843} & \multirow{2}{*}{-4,864} & \multirow{2}{*}{-153} & \multirow{2}{*}{10,961} & \multirow{2}{*}{-7,351} & \multirow{2}{*}{5,157} & \multirow{2}{*}{168} & \multirow{2}{*}{-3,706} & \multirow{2}{*}{1,539} \\
(\texteuro \ per capita) &  &  &  &  &  & &  & &  & &  &  \\
\bottomrule
\end{tabular}
\annote{This table presents the doppelganger output gap per capita in 2007. This measure is obtained by adjusting the real GDP gap for the population size and converting 2011 US dollars into 2011 euro. We use the conversion rate available from the PWT 9.1 for this year ($\approx 0.73$).}
\end{table}

@@

A key assumption to study the impact of a policy intervention is that there is no reverse causality. In our context, this means that countries must not have adopted the euro due to economic considerations. This assumption is plausible because the Eurozone accession was driven mainly by political rather than economic factors \citep{Feldstein1997, Dyson1999, Willett2000, Spolaore2013}.\footnote{This argument holds even for the Greek case which had decided to join the euro before the single currency was a reality. According to the 1998 convergence report from the European Commission, Greece did not join the single currency in 1999 because it had not fulfilled any of the four convergence criteria. Notwithstanding, the decision to join was already made.} The adoption of the Euro was seen more as a political project aimed at deepening European integration; reducing the risks of conflict; and creating a stronger Europe on the global affair \citep{Eichengreen1993, James2012}.

\cite{Feldstein1997} found that the economic impact of joining the EMU would be uncertain, with the possibility of some small gains in trade and investment, accompanied by larger unemployment and inflation. \cite{Fernandez2013} studies the mechanisms through which the adoption of the Euro delayed, rather than advanced, economic reforms in some Eurozone countries.

Moreover, the Euro area is not considered an optimal currency area (OCA), which casts doubt on the strength of the economic motivations behind its adoption. According to \citep{Mundell1961}, OCA have labor mobility, capital mobility, a risk-sharing system such as a fiscal transfer mechanism, and similar business cycles among the member countries. Labor mobility was limited by language and cultural barriers, and there was significant divergence in economic conditions and fiscal policies among the member states. Importantly, by not satisfying the requirements of an Optimum Currency Area, many economists believed that some countries adopting the euro would face economic losses \citep{Jonung2009}.

Even though there are expected gains in terms of real economic growth from adopting a fixed exchange regime, that is not expected for all countries alike; \cite{Arvai2023} provide evidence that countries with less credible monetary policy are the ones growing after pegging their currency to a more credible anchor.

Sections \ref{SS_Implementation} and \ref{SS_Data} discussed the conditions under which the SCM provides suitable estimates of causal effects and this section addresses some of these requirements. To further back the notion that the doppelganger gap is indeed caused by the euro adoption, Sections \ref{SS_timeplacebo} and \ref{SS_spaceplacebo} provide two key placebo experiments. We can be confident that the synthetic control estimator captures the causal effect of an intervention as long as similar magnitudes are not estimated in cases where the intervention did not take place \citep{Born2018}. Finally, Section \ref{SS_stat} discusses the statistical significance of the results.

@@

\begin{figure}[h!]
    \centering
    \caption{In-time Placebo Tests}
    \includegraphics[scale=1.2]{}
    \annote{In each graph, the blue dashed line represents the normalized real GDP for the synthetic country and the black full line represents the series for the actual country. The shaded area corresponds to two standard deviations of the difference between the treated country and the doppelganger prior to the euro accession. The vertical line depicts the placebo treatment period - 1992 for all countries. For all countries, the analysis starts in 1970 and ends in 2007. %The baseline specification has as predictors labor productivity growth, employment share, and GDP shares from PENN Tables 9.1 data. OECD restricted donor pool.
    }
    \label{F_Maastricht}
\end{figure}

@@

The absence of anticipation effects for the remaining countries may be due to two things. First, the key event representing a change for most European citizens was the irrevocable exchange rate fix on 31st December 1998 and the euro launch on the 1st of January 1999. Second, most of these countries had already experienced trade and economic gains from joining the European Union \citep{Campos2018}. Therefore, such effects lie in our pre-treatment sample and thus, are already being considered.\footnote{Add the fact that Italy dropped the exchange rate agreement and floated for a year 93-94. Add that Austria and Finland joined the EU in 1995. Add that }
%Still, these figures should be analysed carefully. For example, the plot for Greece shows some effect that starts even before the placebo treatment of 1992. This anticipated effect seems to be driven by the Greek European Union (EU) accession in 1981 and not from the Euro. 
%Furthermore, estimates for Austria and Finland must be interpreted carefully as these countries entered the EU exactly in 1995 and the placebo treatment effect may be biased.

@@

Second, there might have been spillover effects of the treatment on the donor countries. We address this issue by iteratively re-estimating our baseline estimate for each Eurozone country excluding in each iteration one of the countries with positive weight \citep{Abadie2015}. Moreover, this robustness check also tests the hypothesis that a single country in the donor pool biased the results. If the exclusion of a single unit from the donor pool has a large effect on results--without a discernible change in pre-intervention fit--the estimated results might be caused by other interventions or by particularly idiosyncratic shocks on the outcome of the excluded untreated unit.

We display this robustness check for countries that weighted, at least, 10\% in the construction of, at least, 2 countries' counterfactuals in Appendix \ref{SS_donor_appendix}. This exercise shows that no particular donor country is driving the main conclusions. So, it is unlikely that there were spillover effects of the treatment on the donors or that one specific country in the donor pool is driving the results.\footnote{The robustness of SCM comes exactly from having similar results using different donors, even if the estimated weights change.}
%\footnote{For the sake of brevity, the figures for all the remaining countries are not reported, but they can be provided by the authors upon request. The results remain qualitatively unchanged.} So, it is unlikely that spillover effects nor one specific country in the donor pool are driving the results.

@@

For countries that joined the EU at least 10 years before the Eurozone creation, we re-do the estimates using their accession data as the start of our sample, that is, Ireland (1973), Greece (1981), Portugal, and Spain (1986). From Figure \ref{F_Start}, it is possible to conclude that the results from the baseline analysis are robust.

@@

\section{Unpacking the GDP gap \label{S_Components}}

In this section, we take one step further and inspect the results presented in Figure \ref{F_1} by decomposing the euro accession response of GDP into the response of its components. First, we compute the series for each GDP component for both countries and corresponding counterfactuals. Then, we uncover which components were positively or negatively affected and account for the output gains and losses from the accession.

@@

\textbf{Countries that implemented spending-based austerity measures between 1992 and 1998} Austria, Belgium, Germany, Spain, Finland, France, Italy

\textbf{Countries that did not implement spending-based austerity measures between 1999 and 2007} Austria, Belgium, Spain, Finland, France,

Italy and Portugal did implement strong austerity measures between 2004 and 2006 accounting for more than 3\% of GDP in total for each country.

@@

For most countries, government consumption is one of the components with a positive contribution. On the one hand, this is a puzzling finding given the Eurozone countries were operating under the fiscal rules of the Stability and Growth Pact (SGP) during the post-treatment period.\footnote{Nevertheless, according to \cite{Bofinger2003}, the correlation between the average real growth rate of government consumption and fiscal deficits between 1999 and 2002 was negative, meaning that the low deficit rules did not necessarily imply less government spending.}

On the other hand, before the Euro accession, there were stringent Maastricht convergence criteria which might have restricted government consumption, creating then more fiscal space after 1999. There seems to be indeed a strong negative relationship between the fiscal spending before and after the Euro adoption. Most countries in the sample, like Finland and Spain, implemented spending-based austerity measures after the Maastricht Treaty in line with the Stability and Growth Pact (SGP) \citep{Alesina2019}.\footnote{Countries that implemented spending-based austerity measures between 1992 and 1998 were Austria, Belgium, Germany, Spain, Finland, France, and Italy. Countries that did not implement spending-based austerity measures between 1999 and 2007 were Austria, Belgium, Spain, Finland, and France.} This has likely created fiscal space which together with access to cheaper credit promoted an increase in government consumption beyond what would have happened absent the Euro adoption. Other countries, such as Portugal, that did not consolidate spending before the Euro accession, also did not boost government consumption afterwards.

@@

Table \ref{T_Decomposition_NX} indeed reveals that, except for France and Italy, all countries had a higher trade volume than if they had not adopted the common currency. This result is in accordance with \cite{Baldwin2008} and \cite{Schmitz2011} who argue that the euro has significantly promoted trade in the Eurozone countries. Yet, net exports changed differently across countries. Only Germany and Ireland experienced significant net trade benefits from the euro accession. We corroborate the negative impact on net exports in most cases as documented in \cite{Hope2016}.

@@

% \subsection{What happened after the Maastricht Treaty might help us understand some of the dynamics we see after the Euro adoption.}

% \begin{figure}[h!]
%     \centering
%     \caption{Doppelganger Gaps and GDP Components}
%     \includegraphics[scale=0.8]{Decomposition_1999.jpg}
%     \annote{The dot depicts, for each country, the percent doppelganger gap of output computed as in Equation \ref{equation3}. The stacked bars represent the contribution of each GDP component for these gaps. The values for GDP component sum up the percent doppelganger gap for each treated unit. The values represent the cumulative effect of the euro accession since they are computed for 2007, the last year of the analysis.}
%     \label{F_Decomposition_1999}
% \end{figure}

\section{Did Euro Area Nationals Gain with the Euro Adoption? \label{S_GNI}}
